Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods

### 2023-07-27 paper/main.tex
finished sec 1

\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle
\begin{abstract}
  This paper introduces a novel computational framework for solving
alternating current (AC) optimal power flow (OPF) problems using
graphics processing units (GPUs). While GPUs have demonstrated
remarkable performance in various computing domains, their application
in AC OPF has been limited due to challenges associated with porting
sparse automatic differentiation (AD) and sparse linear solver
routines to GPUs. We aim to address these issues with two key
strategies.  First, we utilize a single-instruction, multiple-data
(SIMD) abstraction of nonlinear programs (NLP). This approach enables
the specification of model equations while preserving their
parallelizable structure, and in turn, facilitates the implementation
of AD routines that can exploit such structure.  Second, we employ a
condensed-space interior-point method (IPM) with an inequality
relaxation strategy. This technique involves relaxing equality
constraints to inequalities and condensing the Karush-Kuhn-Tucker
system into a much smaller positive definite matrix. This strategy
offers the key advantage of being able to factorize the KKT matrix
without numerical pivoting, which has hampered the parallelization of
the algorithm.  By combining these two strategies, we can perform the
majority of operations on GPUs while keeping the data residing in the
device memory only. Comprehensive numerical benchmark results showcase
the substantial computational advantage of our approach. Remarkably,
for solving large-scale AC OPF problems to a moderate accuracy, our
implementations—MadNLP.jl and SIMDiff.jl—running on NVIDIA GPUs
achieve an order of magnitude speedup compared to state-of-the-art
tools running on contemporary CPUs.
\end{abstract}

@@

While derivative evaluation can be generally cheaper than linear
algebra operations, our numerical results show that AD often
constitutes more than half of the total solver time when using
off-the-shelf AD implementations like JuMP.jl or AMPL. Instead of the
general-purpose tools, our method leverages a specialized AD
implementation based on the SIMD abstraction of NLPs. This abstraction
alllows to preserve the parallelizable structure within the model
equations, facilitating efficient derivative evaluations on the
GPU. The AC power flow model is particularly well-suited for this
abstraction as it involves repetitive expressions for each component
type (e.g., buses, lines, generators), and the number of computational
patterns does not increase with the network's size. Numerical results
reported in this paper demonstrate that our proposed strategies can
achieve over 20 times speedup by running on GPU. In comparison to
general AD implementations on CPUs (such as AMPL and JuMP.jl), our
GPU-based method can be approximately 50 times faster.

Linear algebra operations, especially sparse indefinite matrix
factorization, are typically the bottleneck in NLP solution methods.
Parallelizing this operation has been considered to be challenging,
primarily due to the need for numerical pivoting, which is sequential
in nature. However, when the matrix can be factorized without
numerical pivoting, a significant part of the operation can be
parallelized, and the numerical factorization can be efficiently
performed on GPUs. We develop a condensed-space IPM strategy that
allows the use of sparse matrix factorization routines without
numerical pivoting. This strategy relaxes equality constraints by
permitting small violations, which enables expressing the
Karush-Kuhn-Tucker (KKT) system entirely in the primal space through
the condensation procedure. Although this strategy is not new
\cite{nocedal2006numerical}, it has traditionally been considered less
efficient than the standard full-space method due to increased nonzero
entries in the KKT system. However, when implemented on GPUs, it
offers the key advantage of ensuring positive definiteness in the
condensed KKT system. This, in turn, allows for the utilization
of linear solvers with a fixed numerical pivot sequence (known as
refactorization), and an efficient implementation is available in the CUDA library.
Although this method is susceptible to numerical stability issues due to
increased condition number in the KKT system, our results demonstrate
that the solver is robust enough to solve problems with a relative
accuracy of $\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$.

We present numerical benchmark results to showcase the efficiency of
our method, utilizing our implementations: MadNLP.jl and
SIMDiff.jl. The solution of the KKT system is performed using the
external cuSOLVER library. To assess the performance of our method, we
compare it against standard CPU approaches using the data available in
pglib-opf \cite{wachter2006implementation}.  Our benchmark results
demonstrate that our proposed computational framework has significant
potential for accelerating the solution of AC OPF problems, especially
when a moderate accuracy (e.g., $10^{-4}$) is sufficient.
Specifically, when running on NVIDIA GPUs, our method achieves a
notable 4x speedup compared to our solver running on CPU, in the case
of the largest instance. Moreover for the same instance, our approach
surpasses the performance of existing tools (such as Ipopt interfaced
with JuMP.jl) by an order of magnitude.  This finding underscores the
importance of harnessing the computational power of GPUs for tackling
the computatoinal challenges in power systems.

>>>>

\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle
\begin{abstract}
  This paper introduces a novel computational framework for solving
alternating current (AC) optimal power flow (OPF) problems using
graphics processing units (GPUs). While GPUs have demonstrated
remarkable performance in various computing domains, their application
in AC OPF has been limited due to challenges associated with porting
sparse automatic differentiation (AD) and sparse linear solver
routines to GPUs. We aim to address these issues with two key
strategies.  First, we utilize a single-instruction, multiple-data
(SIMD) abstraction of nonlinear programs (NLP). This approach enables
the specification of model equations while preserving their
parallelizable structure, and in turn, facilitates the implementation
of AD routines that can exploit such structure.  Second, we employ a
condensed-space interior-point method (IPM) with an inequality
relaxation strategy. This technique involves relaxing equality
constraints to inequalities and condensing the Karush-Kuhn-Tucker
system into a much smaller positive definite matrix. This strategy
offers the key advantage of being able to factorize the KKT matrix
without numerical pivoting, which in the past has hampered the parallelization of
the algorithm.  By combining these two strategies, we can perform the
majority of operations on GPUs while keeping the data residing in the
device memory only. Comprehensive numerical benchmark results showcase
the substantial computational advantage of our approach. Remarkably,
for solving large-scale AC OPF problems to a moderate accuracy, our
implementations—MadNLP.jl and SIMDiff.jl—running on NVIDIA GPUs
achieve an order of magnitude speedup compared to state-of-the-art
tools running on contemporary CPUs.
\end{abstract}

@@

While derivative evaluation can be generally cheaper than linear
algebra operations, our numerical results show that AD often
constitutes more than half of the total solver time when using
off-the-shelf AD implementations like JuMP.jl or AMPL. Instead of the
general-purpose tools, our method leverages a specialized AD
implementation based on the SIMD abstraction of NLPs. This abstraction
alllows to preserve the parallelizable structure within the model
equations, facilitating efficient derivative evaluations on the
GPU. The AC power flow model is particularly well-suited for this
abstraction as it involves repetitive expressions for each component
type (e.g., buses, lines, generators), and the number of computational
patterns does not increase with the network's size. Numerical results
reported in this paper demonstrate that our proposed strategies can
achieve over 20 times speedup by running on GPU. In comparison to
general AD implementations on CPUs (such as AMPL and JuMP.jl), our
GPU-based  differentiation method can be approximately 50 times faster.

Linear algebra operations, especially sparse indefinite matrix
factorization, are typically the bottleneck in NLP solution methods.
Parallelizing this operation has been considered to be challenging,
primarily due to the need for numerical pivoting, which is sequential
in nature. However, when the matrix can be factorized without
numerical pivoting, a significant part of the operation can be
parallelized, and the numerical factorization can be efficiently
performed on GPUs. We develop a condensed-space IPM strategy that
allows the use of sparse matrix factorization routines without
numerical pivoting. This strategy relaxes equality constraints by
permitting small violations, which enables expressing the
Karush-Kuhn-Tucker (KKT) system entirely in the primal space through
the condensation procedure. Although this strategy is not new
\cite{nocedal2006numerical}, it has traditionally been considered less
efficient than the standard full-space method due to increased nonzero
entries in the KKT system. However, when implemented on GPUs, it
offers the key advantage of ensuring positive definiteness in the
condensed KKT system. This, in turn, allows for the utilization
of linear solvers with a fixed numerical pivot sequence (known as
refactorization), an efficient implementation of which is available in the CUDA library.
Although this method is susceptible to numerical stability issues due to
increased condition number in the KKT system, our results demonstrate
that the solver is robust enough to solve problems with a relative
accuracy of $\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$.

We present numerical benchmark results to showcase the efficiency of
our method, utilizing our implementations: MadNLP.jl and
SIMDiff.jl. The solution of the KKT system is performed using the
external cuSOLVER library. To assess the performance of our method, we
compare it against standard CPU approaches using the data available in
pglib-opf \cite{wachter2006implementation}.  Our benchmark results
demonstrate that our proposed computational framework has significant
potential for accelerating the solution of AC OPF problems, especially
when a moderate accuracy (e.g., $10^{-4}$) is sufficient.
Specifically, when running on NVIDIA GPUs, our method achieves a
notable 4x speedup compared to our solver running on CPU, in the case
of the largest instance. Moreover for the same instance, our approach
surpasses the performance of existing tools (such as Ipopt interfaced
with JuMP.jl) by an order of magnitude.  This finding underscores the
importance of harnessing the power of GPUs for tackling
the computational challenges in power systems.
